---
title: "ITO66-patient"
author: "Thomas W. Battaglia"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.path = "images/")
library(knitr)
library(patchwork)
library(ggsci)
library(ggrepel)
library(DESeq2)
library(readxl)
library(ggplotify)
library(plotly)
library(ggpubr)
library(tidyverse)
source('helpers.R')
```

# Introduction

## Data import & preprocessing
We are going to first import the merged counts tables that is outputted from the Nextflow pipeline. This table already has the raw counts summarised from the alignments and merged into a single matrix.

## Import data
```{r}
# Import CD4 counts
counts = readRDS('../data/ITO66-counts.rda')
head(counts)

# Import CD4 metadata
metadata = readRDS('../data/ITO66-metadata.rda')
head(metadata)

# Import MultiQC results
report = readRDS('../data/ITO66-report.rda')
head(report)
```


----

### Number of mapped reads

```{r}
p1 = report %>% 
  select(Sample, reads_unmapped, reads_mapped) %>% 
  mutate(Sample = str_replace(Sample, '-bamstats', '')) %>% 
  pivot_longer(cols = c("reads_unmapped", "reads_mapped"), names_to = "feature", values_to = "reads") %>% 
    mutate(feature = if_else(feature == "reads_unmapped", "Unmapped", feature)) %>% 
  mutate(feature = if_else(feature == "reads_mapped", "Mapped", feature)) %>% 
  ggplot(aes(x = reorder(Sample, reads), y = reads, fill = feature)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 10) +
  theme(legend.position = "top") +
  scale_fill_brewer(palette = "Set2") +
  xlab("") +
  ylab("Number of reads") +
  ggtitle("Library mapping rate")
p1

### Percent of mapped reads
p2 = report %>% 
  select(Sample, reads_unmapped_percent, reads_mapped_percent ) %>% 
  mutate(Sample = str_replace(Sample, '-bamstats', '')) %>% 
  pivot_longer(cols = c("reads_unmapped_percent", "reads_mapped_percent"), names_to = "feature", values_to = "reads") %>% 
  mutate(reads = reads / 100) %>% 
  mutate(feature = if_else(feature == "reads_unmapped_percent", "Unmapped", feature)) %>% 
  mutate(feature = if_else(feature == "reads_mapped_percent", "Mapped", feature)) %>% 
  ggplot(aes(x = reorder(Sample, reads), y = reads, fill = feature)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 10) +
  theme(legend.position = "top") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent) +
  xlab("") +
  ylab("Percent of reads")
p2

# Plot together
p = (p1 | p2) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
p
ggsave("figures/ITO66-mapping-qc.pdf", p, width = 8, height = 5)
```



----

## Normalize data

```{r}
# Match the sample ID's between samples
counts = counts[,match(metadata$Sample, colnames(counts))]

# Make DESeq2 object
ddsMat <- DESeqDataSetFromMatrix(countData = counts,
                                 colData = metadata,
                                 design = ~Group)

# Get normalize counts
ddsMat <- estimateSizeFactors(ddsMat)

# Filter flow counts
ddsMat.fil = ddsMat %>% 
  filter_low(percentile = 0.05)

# Get normalized counts 
counts.norm.fil <- (counts(ddsMat.fil, normalized = TRUE) + 0.5) %>% 
  as.data.frame() %>% 
  rownames_to_column("oligo_id")

# Create a long table 
counts.norm.fil.long = counts.norm.fil %>% 
  pivot_longer(cols = -oligo_id, names_to = "sampleId", values_to = "norm")
```

---

### Technical replicates
Bcells_alone_1 appears to be too noisy for inclusion as as replicate

```{r}
# Bcell
counts.norm.fil %>% 
  ggplot(., aes(x = log10(Bcells_alone_1), y = log10(Bcells_alone_2))) +
  geom_point(colour = alpha("grey", 0.7)) +
  xlab("Bcell #1") +
  ylab("Bcell #2") +
  theme_classic(base_size = 9) +
  stat_cor(method = "pearson", aes(label = ..r.label..)) +
  stat_smooth(method = "lm", color = "red") 

counts.norm.fil %>% 
  ggplot(., aes(x = log10(Bcells_alone_1), y = log10(Bcells_cntrl))) +
  geom_point(colour = alpha("grey", 0.7)) +
  xlab("Bcell #1") +
  ylab("Bcell-ctrl #1") +
  theme_classic(base_size = 9) +
  stat_cor(method = "pearson", aes(label = ..r.label..)) +
  stat_smooth(method = "lm", color = "red") 

counts.norm.fil %>% 
  ggplot(., aes(x = log10(Bcells_alone_2), y = log10(Bcells_cntrl))) +
  geom_point(colour = alpha("grey", 0.7)) +
  xlab("Bcell #2") +
  ylab("Bcell-ctrl #1") +
  theme_classic(base_size = 9) +
  stat_cor(method = "pearson", aes(label = ..r.label..)) +
  stat_smooth(method = "lm", color = "red") 

# B & T-cells
counts.norm.fil %>% 
  ggplot(., aes(x = log10(Bcells_Tcells_1), y = log10(Bcells_Tcells_2))) +
  geom_point(colour = alpha("grey", 0.7)) +
  xlab("BT cells #1") +
  ylab("BT cells #2") +
  theme_classic(base_size = 9) +
  stat_cor(method = "pearson", aes(label = ..r.label..)) +
  stat_smooth(method = "lm", color = "red") 

counts.norm.fil %>% 
  ggplot(., aes(x = log10(Bcells_Tcells_1), y = log10(Bcells_Tcells_3))) +
  geom_point(colour = alpha("grey", 0.7)) +
  xlab("BT cells #1") +
  ylab("BT cells #3") +
  theme_classic(base_size = 9) +
  stat_cor(method = "pearson", aes(label = ..r.label..)) +
  stat_smooth(method = "lm", color = "red")

counts.norm.fil %>% 
  ggplot(., aes(x = log10(Bcells_Tcells_2), y = log10(Bcells_Tcells_3))) +
  geom_point(colour = alpha("grey", 0.7)) +
  xlab("BT cells #2") +
  ylab("BT cells #3") +
  theme_classic(base_size = 9) +
  stat_cor(method = "pearson", aes(label = ..r.label..)) +
  stat_smooth(method = "lm", color = "red")

```


### Plot normalized histogram
```{r}
counts.norm.fil.long %>% 
  ggplot(aes(x = log10(norm))) +
  geom_histogram(aes(y=..density..), binwidth=0.25, colour="black", fill="white") +
  geom_density(alpha = 0.2, fill = "#FF6666") +
  facet_wrap(sampleId~.) +
  theme_light(base_size = 11) +
  ylab("Density") +
  xlab("Normalized reads per oligo (log10)") 
```


----

## Determine dropout oligo's
```{r}
# Keep samples of interest
ddsMat.fil = ddsMat[,which(colnames(assay(ddsMat)) %in% c("Bcells_alone_2", "Bcells_cntrl", "Bcells_Tcells_2", "Bcells_Tcells_1"))]

# Detect dropouts
dropout.res = compute_stats(ddsMat.fil, trt = "BTcell", ref = "Bcell", fitType = "parametric", lfcThreshold = 0.25, alpha = 0.05) %>% 
  mutate(log2FoldChange = round(log2FoldChange, 2)) %>% 
  mutate(log2FoldChangeShrunken = round(log2FoldChangeShrunken, 2)) 

# Significant oligos
dropout.res %>% 
  filter(padj < 0.05)

# Browse
dropout.res %>% 
  select(gene_id, log2FoldChange, padj, gene_name, gene_id) %>% 
  DT::datatable(extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  ))

# Plot MA-plot
ito66.maplot = dropout.res %>% 
  ggplot(aes(x = log2(baseMean), y = log2FoldChange, label = oligo_id)) +
  geom_point(colour = alpha("grey", 0.7)) +
  geom_point(data = filter(dropout.res, oligo_id %in% c("ABT1_001")), size = 3, colour = "#efc383") +
  geom_point(data = filter(dropout.res, oligo_id %in% c("EEF1A1_001B")), size = 3, colour = "#efc383") +
  geom_text_repel(data = filter(dropout.res, oligo_id %in% c("ABT1_001", "EEF1A1_001B")), box.padding = 0.5, max.overlaps = Inf, size = 2.5) +
  theme_classic() +
  xlim(5, 10) +
  ylim(-3, 2) +
  xlab("Mean counts (log2)") +
  ylab("Fold change log2(B&Tcells/Bcells)")  +
  ggtitle("ITO66")
ggplotly(ito66.maplot)
```




## Session info
```{r}
sessionInfo()
```

